---
title: "exercise"
author: "Manuel"
format: pdf
editor: visual
---

### Aim

Simplify complex community data using and exploring various standard methodology (community similarity, ordination, classification).

We do so to test the effect of human arrival on island communities.

Valuable internet source:\
<https://www.davidzeleny.net/anadat-r/doku.php/en:ordination>

------------------------------------------------------------------------

### Load the data

```{r}
library(here)
```

We start loading the pollen core of our focal island (Tenerife)

```{r}
dat <- read.csv2(here("Data", "Tenerife.csv"))
age <- dat$age # sample ages 
species <- dat[, 2:ncol(dat)] # pollen "species" (taxa) 
rm(dat)

head(species)
```

### Your task:

Explore the data by plotting the pollen counts of any species with time.

```{r}
# Your solution:

```

### Optional task (but nice):

Use a stratigraphic plot to get a better overview over all species and their temporal patterns. Function Stratiplot in library analogue is a possible option. Feel free to use another one (like the ggplot solution applied yesterday).

```{r}
# Your solution:
# install.packages("analogue")
# library(analogue)
# Stratiplot()
```

### Ordination I: Correspondence analysis (CA)

Ordination methods help you to get a quick overview of patterns in community composition. The **cca()** function in the vegan package allows you to run, a correspondence analysis (CA) that analyses community data. The function will run a CA if you only provide a community data matrix (in our case species).

INFO for later: You can also use the function to run a constrained correspondence analysis (CCA) that allows you to explore changes in community composition with environmental gradients. But there are other methods like jSDMs, which are more powerful for this. More information for later: browseVignettes("vegan")

Now lets run the **Correspondence analysis (CA)**. Please take time to look at the visualization and discuss what it shows (and what not). The axes represent gradients in species composition. Species scores and site scores are the coordinates of the species and sites on these axes:

```{r}
ord<-vegan::cca(species)
plot(ord)
```

Eigenvalues can be interpreted as the explained variance (importance) of the different axes. The first axis always explains more than consecutive ones.

```{r}
summary(ord)
```

Function **screeplot()** allows you to visualize the eigenvalues. It shows how the importance of a single axis declines and may help to decide how many axes you would like to look at. We only show the first two axes.

```{r}
screeplot(ord)
```

If you would like to differentiate the pattern of species and sites, this can be done using the argument display, where you can either select species or sites (both in quotation marks).

```{r}
plot(ord, type = "n")
points(ord, display = "sites", cex = 0.8, pch=21, col="red", bg="yellow")
text(ord, display = "species", cex=0.7, col="blue")
```

Now let us add time to the plots to see how species composition changes with time

```{r}
CA1<-ord$CA$u[,1] # coordinates of first axis
CA2<-ord$CA$u[,2] # coordinates of second axis

# Plot adding the ages
plot(ord, type = "n")  # make empty plot
points(ord, display = "sites", cex = 0.8, pch=21, col="red", bg="yellow",type="b")
text(x=CA1+0.1,y=CA2+0.1, cex=0.7,labels=age, col="blue")
```

### Your task:

Now, we want to visualize how species composition on Tenerife has changes with time. To do so, plot the strongest gradient in species composition (first axis) against time. Humans arrived at the island about 2300 years ago. Indicate this event with a vertical line Make the plot nice!

```{r}
# Your solution:

```

**INFO, not for now:** The CA axes are the most important gradients in species composition in the data. You can also test how those are reflecting environmental gradients. In vegan, this is nicely implemented in function envfit()

### Optional task (but nice):

There is a second island data set available. First, check where it is. Then Explore how pollencommunity composition has changed on that island.

```{r}
# Your solution:

```

# STOP HERE !

### Similarity in species composition

vegdist allows you to calculate similarity in species composition between communities. It transfers the community matrix into a distance object. In our case, it compares each level in the core with all the others. The standard setting calculates Bray–Curtis dissimilarity. Have a look at the original community matrix as well as the distance object (dist):

```{r}
head(species) # community matrix
dist<-vegan::vegdist(species)
head(dist) # distance object
```

### Ordination 2: Non-metric multidimensional scaling

```{r}
nmds<-vegan::metaMDS(dist,  trace = TRUE) # runs the NMDS

nmds$points # extract coordinates

# NMDS 2D figure:
plot(nmds, type = "n")
points(nmds, display = "sites", cex = 0.8, pch=21, col="red", bg="yellow",type="b")
text(nmds, labels = age, cex=0.7, col="blue")
```

### Cluster analysis

Function **hclust** is one way to run a classification, allowing various different methods (see help). There are multiple other packages with alternatives. You decide on the method and the number of classes. Please try 2-3 different methods to see the sensitivity:

```{r}
tree<-hclust(dist,method = "complete") # classification

plot(tree) 

rect.hclust(tree, 5, border="red") # cut the tree forming 5 classes

# try other methods and vary the number of classes:

```

FINALLY: Use the function cutree to extract the class for each layer in our core. Add that information to the plot displaying the first ordination axis against time by changing the point color according to the classification result e.g. col=cutree(tree,k=2)

```{r}
# Your solution:

```

### Optional task:

Now select and load a core from neotoma as shown by Ondřej yesterday. Use the ordination to relate the major changes in species composition with time. And check if a major war (like the 30 years war in Europe), the plague or any other possible driver is visible in pollen composition of the landscape. Find a way to test/visualize the effect across 1000 cores and write the paper up.

```{r}
# Your solution:

```

### 
